Calibration tests beyond classification

### 2022-05-30 paper/main.tex
Correct comments about kernel choice

By definition, so-called characteristic kernels guarantee that
$\mathrm{KCE} = 0$ if and only if the distributions of $(P_X,Y)$
and $(P_X,Z_X)$ are equal~\citep{Fukumizu2004,Fukumizu2008}. Many common kernels
such as the Gaussian and Laplacian kernel on $\mathbb{R}^d$ are
characteristic~\citep{Fukumizu2008}.\footnote{For a general discussion about
characteristic kernels and their relation to universal kernels we refer to the 
paper by \citet{Sriperumbudur2011}.} \Citet[Theorem~4]{Szabo2018} showed that a
tensor product kernel $k_{\mathcal{P}} \otimes k_{\mathcal{Y}}$ is characteristic if $k_{\mathcal{P}}$ and
$k_{\mathcal{Y}}$ are characteristic, continuous, bounded, and translation-invariant kernels
on $\mathbb{R}^d$, but the implication does not hold for general characteristic
kernels~\citep[Example~1]{Szabo2018}. For calibration evaluation, however,
it is sufficient to be able to distinguish between the conditional distributions
$\Prob(Y|P_X)$ and $\Prob(Z_X|P_X) = P_X$.
Therefore, in contrast to the regular MMD setting, it
is \emph{sufficient that kernel $k_{\mathcal{Y}}$ is characteristic and kernel $k_{\mathcal{P}}$ is
non-zero almost surely}, to guarantee that $\mathrm{KCE} = 0$ if and only if model $P$ is calibrated.
Thus it is suggestive to construct kernels on general spaces of predicted distributions 
as
\begin{equation}\label{eq:kernelp}
k_{\mathcal{P}}(p, p') = \exp{\big(- \lambda d^\nu_{\mathcal{P}}(p, p') \big)},
\end{equation}
where $d_{\mathcal{P}}(\cdot, \cdot)$ is a metric on $\mathcal{P}$ and $\nu, \lambda > 0$ 
are kernel hyperparameters. The Wasserstein distance is a widely used metric for
distributions from optimal transport theory that allows to lift a ground
metric on the target space and possesses many important
properties~\citep[see, e.g.,][Chapter~2.4]{Peyre2018}. In general, however,
it does not lead to valid kernels $k_{\mathcal{P}}$, apart from the notable exception
of elliptically contoured distributions such as normal and Laplace
distributions~\citep[Chapter~8.3]{Peyre2018}.

@@

@@

@@

As discussed in \cref{sec:kernel_choice}, if kernel $k$ is characteristic,
then the kernel calibration error $\mathrm{KCE}_k$ of model $P$ is zero
if and only if $P$ is calibrated. Unfortunately, as shown by
\citet[Example~1]{Szabo2018}, even if $k_{\mathcal{P}}$ and $k_{\mathcal{Y}}$ are characteristic,
the tensor product kernel $k = k_{\mathcal{P}} \otimes k_{\mathcal{Y}}$ might not be characteristic.
However, when analyzing calibration, it is sufficient to be able to
distinguish distributions for which the conditional distributions $\Prob(Y|P_X)$
and $\Prob(Z_X|P_X) = P_X$ are not equal almost surely. Thus it is sufficient
if $k_{\mathcal{Y}}$ is characteristic and $k_{\mathcal{P}}$ is non-zero almost surely.

Many common kernels such as the Gaussian and Laplacian kernel on
$\mathbb{R}^d$ are characteristic and can therefore be chosen as kernel
$k_{\mathcal{Y}}$ for real-valued target spaces. The choice of $k_{\mathcal{P}}$ might be less obvious
since $\mathcal{P}$ is a space of probability distributions. Intuitively
one might want to use kernels of the form
\begin{equation}\label{eq:kernelform}
    k_{\mathcal{P}}\big(p, p'\big) = \exp{\big(- \lambda d^\nu_{\mathcal{P}}(p, p')\big)},
\end{equation}
where $d_{\mathcal{P}} \colon \mathcal{P} \times \mathcal{P} \to \mathbb{R}$ is a
metric on $\mathcal{P}$ and $\nu, \lambda > 0$ are kernel hyperparameters.
Kernels of this form would be a generalization of the Gaussian and Laplacian
kernel, and would clearly be non-zero almost surely.

@@

@@

Hence we see that a Gaussian kernel
\begin{equation*}
    k_{\mathcal{Y}}(y, y') = \exp{\big(-\gamma {\|y - y'\|}_2^2\big)}
\end{equation*}
with inverse length scale $\gamma > 0$ on the space of
targets $\mathcal{Y} = \mathbb{R}^d$
allows us to compute $\Exp_{Z_x \sim P_x} k_{\mathcal{Y}}(Z_x, y)$ and
$\Exp_{Z_x \sim P_x, Z_{x'} \sim P_{x'}} k_{\mathcal{Y}}(Z_x, Z_{x'})$
analytically. Moreover, the Gaussian kernel is characteristic on
$\mathbb{R}^d$~\citep{Fukumizu2008}. Hence, as discussed above,
by choosing a kernel $k_{\mathcal{P}}$ that is non-zero almost surely we can
guarantee that $\mathrm{KCE}_k = 0$ if and only if model $P$ is
calibrated.

On the space of normal distributions, the 2-Wasserstein distance with
respect to the Euclidean distance between
$P_x = \mathcal{N}(\mu_x, \Sigma_x)$ and
$P_{x'} = \mathcal{N}(\mu_{x'}, \Sigma_{x'})$ is given by
\begin{equation*}
    W^2_2\big(P_x, P_{x'}\big)
    = {\|\mu_x - \mu_{x'}\|}_2^2 + \mathrm{Tr}\bigg(\Sigma_x + \Sigma_{x'} - 2{\Big({\Sigma_{x'}}^{1/2} \Sigma_x {\Sigma_{x'}}^{1/2}\Big)}^{1/2}\bigg),
\end{equation*}
which can be simplified to
\begin{equation*}
    W^2_2\big(P_x, P_{x'}\big)
    = {\big\|\mu_x - \mu_{x'}\big\|}_2^2 + {\Big\|\Sigma_{x}^{1/2} - \Sigma_{x'}^{1/2}\Big\|}_{\mathrm{Frob}}^2,
\end{equation*}
if $\Sigma_x \Sigma_{x'} = \Sigma_{x'} \Sigma_x$. This shows that the
2-Wasserstein distance is a Hilbertian metric on the space of normal
distributions. Hence as discussed above, the choice
\begin{equation*}
    k_{\mathcal{P}}\big(P_x, P_{x'}\big) = \exp{\big( - \lambda W_2^\nu(P_x, P_{x'})\big)}
\end{equation*}
yields a valid kernel for all $\lambda > 0$ and $\nu \in (0, 2]$.

Thus for all $\lambda, \gamma > 0$ and $\nu \in (0, 2]$
\begin{equation*}
    k\big((p, y), (p', y')\big) = \exp{\big(-\lambda W^\nu_2(p, p')\big)} \exp{\Big(-\gamma {\|y-y'\|}^2_2\Big)}
\end{equation*}
is a valid kernel on the product space $\mathcal{P} \times \mathcal{Y}$
of normal distributions on $\mathbb{R}^d$ and $\mathbb{R}^d$ that allows to evaluate
$h\big((p, y), (p', y')\big)$ analytically and guarantees that $\mathrm{KCE}_k = 0$
if and only if model $P$ is calibrated.

@@

Since the Laplace distribution is an elliptically contoured
distribution, we know from \citet[Corollary~2]{Gelbrich1990} that
the 2-Wasserstein distance with respect to the Euclidean distance between
$P_x = \mathcal{L}(\mu_x, \beta_x)$ and
$P_{x'} = \mathcal{L}(\mu_{x'}, \beta_{x'})$ can be
computed in closed form and is given by
\begin{equation*}
    W^2_2\big(P_x, P_{x'}\big) = {(\mu_x - \mu_{x'})}^2 + 2{(\beta_x - \beta_{x'})}^2.
\end{equation*}
Thus we see that the 2-Wasserstein distance is also a Hilbertian metric
on the space of Laplace distributions, and hence
\begin{equation*}
    k_{\mathcal{P}}\big(P_x, P_{x'}\big) = \exp{\big(- \lambda W_2^\nu(P_x, P_{x'})\big)}
\end{equation*}
is a valid kernel for $0 < \nu \leq 2$ and all $\lambda > 0$.

Therefore, as discussed above, for all $\lambda, \gamma > 0$ and $\nu \in (0, 2]$
\begin{equation*}
    k\big((p, y), (p', y')\big) = \exp{\big(-\lambda W^\nu_2(p, p')\big)} \exp{\big(-\gamma |y-y'|\big)}
\end{equation*}
is a valid kernel on the product space $\mathcal{P} \times \mathcal{Y}$
of Laplace distributions and $\mathbb{R}$ that allows to evaluate
$h\big((p, y), (p', y')\big)$ analytically and guarantees that $\mathrm{KCE}_k = 0$
if and only if model $P$ is calibrated.

@@

Thus taken together, if $k_{\mathcal{Y}}$ is a characteristic kernel on the
target space $\mathcal{Y}$ and $d(\cdot, \cdot)$ is a Hilbertian
metric on the space of mixture components, then for all $s \in [1, \infty)$, $\lambda > 0$,
and $\nu \in (0, 2]$
\begin{equation*}
    k\big((p, y), (p', y')\big) = \exp{\big(-\lambda \mathrm{MW}_s^\nu(p, p')\big)} k_{\mathcal{Y}}(y, y')
\end{equation*}
is a valid kernel on the product space $\mathcal{P} \times \mathcal{Y}$
of mixture distributions and targets that allows to evaluate $h\big((p, y), (p', y')\big)$
analytically and guarantees that $\mathrm{KCE}_k = 0$ if and only if
model $P$ is calibrated.

>>>>

So-called characteristic kernels guarantee that $\mathrm{KCE} = 0$ if and only if the distributions of $(P_X,Y)$ and $(P_X,Z_X)$ are equal, i.e., if model $P$ is calibrated~\citep{Fukumizu2004,Fukumizu2008}.
Many common kernels such as the Gaussian and Laplacian kernel on $\mathbb{R}^d$ are characteristic~\citep{Fukumizu2008}.%
\footnote{For a general discussion about characteristic kernels and their relation to universal kernels we refer to the paper by \citet{Sriperumbudur2011}.}
For kernels $k_{\mathcal{P}}$ and $k_{\mathcal{Y}}$ their characteristic property is necessary but generally not sufficient for the tensor product kernel $k_{\mathcal{P}} \otimes k_{\mathcal{Y}}$ to be characteristic~\citep[Example~1]{Szabo2018}.
However, if $k_{\mathcal{P}}$ and $k_{\mathcal{Y}}$ are characteristic, continuous, bounded, and translation-invariant kernels on $\mathbb{R}^{d_i}$ ($i=1,2$), then $k_{\mathcal{P}} \otimes k_{\mathcal{Y}}$ is characteristic~\citep[Theorem~4]{Szabo2018}.
We use this property to construct characteristic tensor product kernels for regression problems.
More generally, if $k_{\mathcal{P}}$ and $k_{\mathcal{Y}}$ are universal kernels on locally compact Polish spaces, then $k_{\mathcal{P}} \otimes k_{\mathcal{Y}}$ is universal and hence characteristic~\citep[Theorem~5]{Szabo2018}.
In classification even the reverse implication holds, and $k_{\mathcal{P}} \otimes k_{\mathcal{Y}}$ is characteristic if and only if $k_{\mathcal{P}}$ and $k_{\mathcal{Y}}$ are universal~\citep[Corollary~3.15]{Steinwart2021}.%

It is suggestive to construct kernels on general spaces of predicted distributions as
\begin{equation}\label{eq:kernelp}
k_{\mathcal{P}}(p, p') = \exp{\big(- \lambda d^\nu_{\mathcal{P}}(p, p') \big)},
\end{equation}
where $d_{\mathcal{P}}(\cdot, \cdot)$ is a metric on $\mathcal{P}$ and $\nu, \lambda > 0$ 
are kernel hyperparameters. The Wasserstein distance is a widely used metric for
distributions from optimal transport theory that allows to lift a ground
metric on the target space and possesses many important
properties~\citep[see, e.g.,][Chapter~2.4]{Peyre2018}. In general, however,
it does not lead to valid kernels $k_{\mathcal{P}}$, apart from the notable exception
of elliptically contoured distributions such as normal and Laplace
distributions~\citep[Chapter~8.3]{Peyre2018}.

@@

\subsection{Empirical properties and computational efficiency}\label{app:efficiency}

@@

\subsection{Friedman 1 regression problem}\label{app:friedman}

@@

As discussed in \cref{sec:kernel_choice}, if kernel $k$ is characteristic,
then the kernel calibration error $\mathrm{KCE}_k$ of model $P$ is zero
if and only if $P$ is calibrated.
Unfortunately, as shown by~\citet[Example~1]{Szabo2018}, for kernels $k_{\mathcal{P}}$ and $k_{\mathcal{Y}}$ their characteristic property is necessary but generally not sufficient for the tensor product kernel $k = k_{\mathcal{P}} \otimes k_{\mathcal{Y}}$ to be characteristic.
Hence in general stronger requirements on $k_{\mathcal{P}}$ and $k_{\mathcal{Y}}$ are needed.
For instance, if $k_{\mathcal{P}}$ and $k_{\mathcal{Y}}$ are characteristic, continuous, bounded, and translation-invariant kernels on $\mathbb{R}^{d_i}$ ($i = 1,2$), then $k_{\mathcal{P}} \otimes k_{\mathcal{Y}}$ is characteristic~\citep[Theorem~4]{Szabo2018}.
More generally, if $k_{\mathcal{P}}$ and $k_{\mathcal{Y}}$ are universal kernels on locally compact Polish spaces, then $k_{\mathcal{P}} \otimes k_{\mathcal{Y}}$ is characteristic~\citep[Theorem~5]{Szabo2018}.
In classification even the reverse implication holds, and $k_{\mathcal{P}} \otimes k_{\mathcal{Y}}$ is characteristic if and only if $k_{\mathcal{P}}$ and $k_{\mathcal{Y}}$ are universal~\citep[Corollary~3.15]{Steinwart2021}.

Many common kernels such as the Gaussian and Laplacian kernel on $\mathbb{R}^d$ are universal (and characteristic), and thus are an appropriate choice for kernel $k_{\mathcal{Y}}$ for real-valued target spaces.
In classification, kernel $k_{\mathcal{Y}}$ is universal if and only if it is strictly positive definite~\citep[Section~3.3]{Sriperumbudur2011} (note that target space $\mathcal{Y}$ is finite).

The choice of $k_{\mathcal{P}}$ might be less obvious
since $\mathcal{P}$ is a space of probability distributions. Intuitively
one might want to generalize Gaussian and Laplacian kernels and use kernels of the form
\begin{equation}\label{eq:kernelform}
    k_{\mathcal{P}}\big(p, p'\big) = \exp{\big(- \lambda d^\nu_{\mathcal{P}}(p, p')\big)},
\end{equation}
where $d_{\mathcal{P}} \colon \mathcal{P} \times \mathcal{P} \to \mathbb{R}$ is a
metric on $\mathcal{P}$ and $\nu, \lambda > 0$ are kernel hyperparameters.

@@

Non-constant radial kernels on $\mathbb{R}^d$ are universal~\citep[Theorem~17]{Micchelli2006}, and hence, with a similar reasoning as in the proof of \citet[Theorem~2.2]{Christmann2010}, $k_{\mathcal{P}}$ is universal on the metric space $(\mathcal{P}, d_{\mathcal{P}})$ for all $\lambda > 0$ and $\nu \in (0, 2]$ if $\phi$ is an injective embedding and $H = \mathbb{R}^d$ for some $d$.
More generally, if $\phi$ is injective and $H$ is a separable Hilbert space, $k_{\mathcal{P}}$ is known to be universal on $(\mathcal{P}, d_{\mathcal{P}})$ for $\lambda > 0$ and $\nu = 2$~\citep[Theorem~2.2]{Christmann2010}.
Note that in both cases it is required that $\phi$ is continuous.
Continuity of $\phi$ is implied by the choice of space $(\mathcal{P}, d_{\mathcal{P}})$ but has to be shown separately if $\mathcal{P}$ is equipped with a different metric.

In many machine learning applications such a mapping $\phi$ arises naturally from the parameterization of the predicted distributions.
For instance, if a regression model predicts normal distributions of real targets, i.e., if $\mathcal{Y} = \mathbb{R}$ and $\mathcal{P} = \{ \mathcal{N}(\mu, \sigma^2) \colon \mu \in \mathbb{R}, \sigma \in \mathbb{R}_{\geq 0} \}$, one may use $H = \mathbb{R}^2$ and define
\begin{equation*}
    \phi\big(\mathcal{N}(\mu, \sigma^2)\big) = [\mu, \sigma]^\mathsf{T}.
\end{equation*}

@@

Hence we see that a Gaussian kernel
\begin{equation*}
    k_{\mathcal{Y}}(y, y') = \exp{\big(-\gamma {\|y - y'\|}_2^2\big)}
\end{equation*}
with inverse length scale $\gamma > 0$ on the space of
targets $\mathcal{Y} = \mathbb{R}^d$
allows us to compute $\Exp_{Z_x \sim P_x} k_{\mathcal{Y}}(Z_x, y)$ and
$\Exp_{Z_x \sim P_x, Z_{x'} \sim P_{x'}} k_{\mathcal{Y}}(Z_x, Z_{x'})$
analytically. Moreover, the Gaussian kernel is universal and characteristic on $\mathbb{R}^d$~\citep{Micchelli2006,Fukumizu2008}. Hence, as discussed above,
by choosing an universal kernel $k_{\mathcal{P}}$ we can
guarantee that $\mathrm{KCE}_k = 0$ if and only if model $P$ is
calibrated.

On the space of normal distributions, the 2-Wasserstein distance with
respect to the Euclidean distance between
$P_x = \mathcal{N}(\mu_x, \Sigma_x)$ and
$P_{x'} = \mathcal{N}(\mu_{x'}, \Sigma_{x'})$ is given by
\begin{equation*}
    W^2_2\big(P_x, P_{x'}\big)
    = {\|\mu_x - \mu_{x'}\|}_2^2 + \mathrm{Tr}\bigg(\Sigma_x + \Sigma_{x'} - 2{\Big({\Sigma_{x'}}^{1/2} \Sigma_x {\Sigma_{x'}}^{1/2}\Big)}^{1/2}\bigg).
\end{equation*}
If $\Sigma_x \Sigma_{x'} = \Sigma_{x'} \Sigma_x$, such as for univariate normal distributions (\cref{app:ols,app:friedman}) or normal distributions with diagonal covariance matrices (\cref{app:efficiency}), this can be simplified to
\begin{equation*}
    W^2_2\big(P_x, P_{x'}\big)
    = {\big\|\mu_x - \mu_{x'}\big\|}_2^2 + {\Big\|\Sigma_{x}^{1/2} - \Sigma_{x'}^{1/2}\Big\|}_{\mathrm{Frob}}^2.
\end{equation*}
In this case the 2-Wasserstein distance is a Hilbertian metric on the space of normal distributions via the injective embedding
\begin{equation*}
    \phi\big(\mathcal{N}(\mu, \Sigma)\big) \coloneqq \big[\mu_1, \ldots, \mu_d, (\Sigma^{1/2})_{1,1}, (\Sigma^{1/2})_{1,2}, \ldots, (\Sigma^{1/2})_{d,d}\big]^\mathsf{T} \in \mathbb{R}^{d(d + 1)}.
\end{equation*}
Hence as discussed above, if the covariance matrices commute, the choice
\begin{equation*}
    k_{\mathcal{P}}\big(P_x, P_{x'}\big) = \exp{\big( - \lambda W_2^\nu(P_x, P_{x'})\big)}
\end{equation*}
yields a valid kernel for all $\lambda > 0$ and $\nu \in (0, 2]$, and is universal on $(\mathcal{P}, W_2)$ for all such parameter choices.

Thus for all $\lambda, \gamma > 0$ and $\nu \in (0, 2]$
\begin{equation*}
    k\big((p, y), (p', y')\big) = \exp{\big(-\lambda W^\nu_2(p, p')\big)} \exp{\Big(-\gamma {\|y-y'\|}^2_2\Big)}
\end{equation*}
is a valid and characteristic kernel on the product space $(\mathcal{P}, W_2) \times (\mathcal{Y}, \|.\|_2)$ where $\mathcal{Y} \subset \mathbb{R}^d$ and $\mathcal{P}$ is a space of normal distributions on $\mathbb{R}^d$ with commuting covariance matrices.
For these kernels $k$, $h\big((p, y), (p', y')\big)$ can be evaluated analytically and it is guaranteed that $\mathrm{KCE}_k = 0$ if and only if model $P$ is calibrated.

@@

Since the Laplace distribution is an elliptically contoured
distribution, we know from \citet[Corollary~2]{Gelbrich1990} that
the 2-Wasserstein distance with respect to the Euclidean distance between
$P_x = \mathcal{L}(\mu_x, \beta_x)$ and
$P_{x'} = \mathcal{L}(\mu_{x'}, \beta_{x'})$ can be
computed in closed form and is given by
\begin{equation*}
    W^2_2\big(P_x, P_{x'}\big) = {(\mu_x - \mu_{x'})}^2 + 2{(\beta_x - \beta_{x'})}^2.
\end{equation*}
Thus we see that the 2-Wasserstein distance is a Hilbertian metric on the space of Laplace distributions via the embedding
\begin{equation*}
    \phi\big(\mathcal{L}(\mu, \beta)\big) := [\mu, \sqrt{2}\beta]^\mathsf{T}.
\end{equation*}
Hence, as discussed above,
\begin{equation*}
    k_{\mathcal{P}}\big(P_x, P_{x'}\big) = \exp{\big(- \lambda W_2^\nu(P_x, P_{x'})\big)}
\end{equation*}
is a valid and universal kernel on $(\mathcal{P}, W_2)$ for $0 < \nu \leq 2$ and all $\lambda > 0$.

Therefore for all $\lambda, \gamma > 0$ and $\nu \in (0, 2]$
\begin{equation*}
    k\big((p, y), (p', y')\big) = \exp{\big(-\lambda W^\nu_2(p, p')\big)} \exp{\big(-\gamma |y-y'|\big)}
\end{equation*}
is a valid and characteristic kernel on the product space $(\mathcal{P}, W_2) \times (\mathcal{Y}, |.|)$ where $\mathcal{Y} \subset \mathbb{R}$ and $\mathcal{P}$ is a space of Laplace distributions.
For these kernels $k$, $h\big((p, y), (p', y')\big)$ can be evaluated analytically and guarantees that $\mathrm{KCE}_k = 0$ if and only if model $P$ is calibrated.

@@

Thus taken together, if $k_{\mathcal{Y}}$ is a kernel on the
target space $\mathcal{Y}$ and $d(\cdot, \cdot)$ is a Hilbertian
metric on the space of mixture components, then for all $s \in [1, \infty)$, $\lambda > 0$,
and $\nu \in (0, 2]$
\begin{equation*}
    k\big((p, y), (p', y')\big) = \exp{\big(-\lambda \mathrm{MW}_s^\nu(p, p')\big)} k_{\mathcal{Y}}(y, y')
\end{equation*}
is a valid kernel on the product space $\mathcal{P} \times \mathcal{Y}$
of mixture distributions and targets that allows to evaluate $h\big((p, y), (p', y')\big)$
analytically.
Moreover, if $k_\mathcal{Y}$ and $k_\mathcal{P}$ are universal kernels, then $k$ is characteristic and hence $\mathrm{KCE}_k = 0$ if and only if model $P$ is calibrated.
